---
title: "DSI Visualization - Assignment 3"
author: "Eman Nishat"
date: "2025-05-11"
output: html_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 12, fig.height = 15)
```

```{r}
# load libraries
library(dplyr)
library(ggplot2)
library(ggpubr)
```

In this report, we will use data from the Ontario Data Catalouge on the Career College Key Performance Indicators to determine the average graduate rate by college in Ontario. 

```{r}
# laod dataset
df <- read.csv("overall_2022_kpi_report_for_publication_full_en.csv")
# check dataset
head(df) 
# check column/variable names
names(df)
# check variable types
summary(df)
```
```{r}
# rename variables of interest
colnames(df)[c(2,11)] <- c("COLLEGE_NAME", "GRAD_RATE")
# check that variables were renamed
names(df)
```
```{r}
# set grad rate as numeric 

## first remove % sign
df$GRAD_RATE <- gsub("%", "", df$GRAD_RATE)
## set as numeric
df$GRAD_RATE <- as.numeric(df$GRAD_RATE)
```

```{r}
# remove rows with NA for GRAD_RATE, and only keep columns of interest (graduate rate and college name)
df_sub <- df[!is.na(df$GRAD_RATE), c(2,11)]
```


```{r}
# calculate average graduation rate by college
college_avg <- df_sub %>%
  # multiple rows for each college due to multiple programs, so we're averaging across those rows
  group_by(COLLEGE_NAME) %>%
  summarize(avg_grad_rate = mean(GRAD_RATE, na.rm = TRUE)) %>%
  # arrange in descending order
  arrange(desc(avg_grad_rate))
```

## Top Colleges

```{r}
# select top 10 colleges with highest graduation rate for readability
top10 <- college_avg %>% 
  slice_max(avg_grad_rate, n = 10)

# creating a bar chart 
ggplot(top10, aes(x = reorder(COLLEGE_NAME, avg_grad_rate), y = avg_grad_rate)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(
    title = "Top 10 Colleges in Ontario by Average Graduation Rate",
    x = "College",
    y = "Average Graduation Rate (%)"
  ) +
  # for plot aesthetics, keep only border and remove any gridlines
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # center the title
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    # rotate the x-axis labels 90 degrees for readability 
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

```

Many colleges have 100$ graduation rate. We want to see other top colleges, with those with perfect rates removed.

```{r}
# filter out colleges with 100% average graduation rate
college_top10 <- college_avg %>%
  # select those with ratings less than 100
  filter(avg_grad_rate < 100) %>%
  # arrange in descending order and select top 10
  arrange(desc(avg_grad_rate)) %>%
  slice_head(n = 10)

top10_plot <- ggplot(college_top10, aes(x = COLLEGE_NAME, y = avg_grad_rate)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(
    title = "Top 10 Ontario Colleges by Graduation Rate",
    x = "College",
    y = "Average Graduation Rate (%)"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )
top10_plot

```

## Botton 10 Colleges

We want to see which colleges have a low graduation rate.

```{r}
# some colleges with have a 0% average graduation rate, we want to exclude these
college_bottom10 <- college_avg %>%
  filter(avg_grad_rate > 0) %>%
  arrange(avg_grad_rate) %>%
  slice_head(n = 10)

# Create bar chart
bottom10_plot <- ggplot(college_bottom10, aes(x = COLLEGE_NAME, y = avg_grad_rate)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(
    title = "Bottom 10 Colleges by Graduation Rate",
    x = "College",
    y = "Average Graduation Rate (%)"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )
bottom10_plot

```

```{r}
# display plots together
combined_plot <- ggarrange(top10_plot, bottom10_plot, ncol = 2, nrow = 1)
# save plot
ggsave("r_barchart.png", plot = combined_plot, width = 12, height = 15, dpi = 300)
```

